---
title: "Assignment 3"
author: "Kimberly Juarez-Rico"
date: "12/5/2021"
output: html_document
---

The purpose of this assignment was to conduct an equity analysis of a population
to better understand if there are any disparities within the selected 
population. 

I chose to use all of the counties in the Bay Area for my analysis. 

Part 1: The first part of this project was to use Public Use Microsample Data
(PUMS) data to pick a factor between a group in the population 
and understand if there are any trends in that factor for that group. 
I chose to analyze Hispanic/Latinx households and see if there was any 
correlation between the number of languages spoken in the household and if 
the household identifies as Hispanic/Latinx. I chose to analyze this factor 
first in order to complete the second part of the project.

```{r setup, include = F}
knitr::opts_chunk$set(warning = F, message = F, echo = F, eval = T)
```

To begin the analysis, I loaded necessary packages:

```{r}
library(tidyverse)
library(sf)
library(tigris)
library(leaflet)
library(censusapi)
library(tidycensus)
library(plotly)
library(gtools)
library(corrplot)
```

Downloaded PUMS data for the 5 year data of 2014-2018:

```{r}
census_api_key("4f290b276a2a5526b13d917c76884e70b99ba725")

pums_vars_2018 <- 
  pums_variables %>%
  filter(year == 2018, survey == "acs5")
```

Refined the PUMS data so that California PUMS Data was divided into the Bay
Area Counties:

```{r}
ca_pums <- get_pums(
  variables = c(
    "PUMA",
    "LANX",
    "HISP"
  ),
  state = "CA",
  year = 2018,
  survey = "acs5",
  recode = T
)

ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_counties <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME %in% bay_county_names)

bay_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_set_geometry(NULL) %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()

bay_pums <-
  ca_pums %>% 
  filter(PUMA %in% bay_pumas$PUMACE10)
```

I then created a binary regression analysis where if the household is hispanic 
and speaks another language other an English, the value is 1 and if they do
not speak another language, the value is 0:

```{r}
bay_pums_regression <-
  bay_pums %>% 
  filter(SPORDER == 1) %>% 
  mutate(
  HISP_label = HISP_label %>% as.character(),
    hispanic = ifelse(
      is.na(HISP_label),
      0,
      1
    ),
   multiple_lang = ifelse(
      LANX_label == "Yes, speaks another language",
      0,
      1
    )
  )
```

I plotted the data into a plot that shows a regression analysis between the 
number of percentage of households that speak another language and are hispanic:

```{r}
ggplot(
  data = bay_pums_regression,
  aes(
      x = hispanic,
      y = multiple_lang
    )
) +
  geom_point() +
  geom_smooth(method = "lm")
```

I also converted the information to a correlation plot that shows what 
percentage of households that identify as hispanic also speak another language.
I found that there was a 100% correlation between households identifying as 
hispanic and speak an additional language other than English. This result is
not surprising since the Bay Area has immigrants from many Hispanic and Latin
American countries that speak another language.

```{r}
correlationplot <- bay_pums_regression %>% 
  select(
    hispanic, 
    multiple_lang
  ) %>% 
  cor()

corrplot(
  correlationplot, 
  method = "number",
  type = "upper"
)
```

While this correlation is interesting, part of the report 2 takes the analysis 
a step further by adding income as another variable to analyze. 

Part 2: The second part of the report involves using American Communities Survey
(ACS) to create a multiple regression analysis of households that
speak Spanish, can speak English well, and income level is over $100K a year in
order to better understand if language barriers in the Hispanic and Latinx 
community affect income. 

First I started by downloading the necessary data for 5 years in 2018:

```{r}
Sys.setenv(install = TRUE, CENSUS_KEY="4f290b276a2a5526b13d917c76884e70b99ba725")

acs_vars_2018_5yr <-
  listCensusMetadata(
    name = "2018/acs/acs5",
    type = "variables"
  )
```

Data sets used at the tract level and for each Bay Area County:
B19001: Household income in the past 12 months (in 2018 inflation-adjusted 
dollars) (Hispanic OR Latino Householder)
B16006: Language spoken at home by ability to speak English for the population 
5 years and over (Hispanic OR Latino)

The ability to speak English is defined as speaking English "well" and "very 
well" to align with the categorizations the ACS communities survey uses. 

The transmute function then calculates the following:
1. The total percent of hispanic people in the Bay Area
2. The total percent of hispanic people in the Bay Area that have a household 
income of  over $100K
3. The total percent of hispanic people in the Bay Area that have a household 
income under $100k
4. The total percent of hispanic people in the Bay Area that speak English well
or very well 
5. The total percent of hispanic people in the Bay Area that speak English not
well or not at all
Columns then filter this information into separate columns for each person
accounted in the survey. 

```{r}
bay_multiple_tract <- 
  getCensus(
    name = "acs/acs5",
    vintage = 2018,
    region = "tract:*",
    regionin = "state:06+county:001,013,041,055,075,081,085,095,097",
    vars = c(
      "B19001I_001E",
      "B19001_001E",
      "B19001_002E",
      "B19001_003E",
      "B19001_004E",
      "B19001_005E",
      "B19001_006E",
      "B19001_007E",
      "B19001_008E",
      "B19001_009E",
      "B19001_010E",
      "B19001_011E",
      "B19001_012E",
      "B19001_013E",
      "B19001_014E",
      "B19001_015E",
      "B19001_016E",
      "B19001_017E",
      "B16006_001E",
      "B16006_004E",
      "B16006_005E",
      "B16006_006E",
      "B16006_007E"
    )
  ) %>% 
  transmute(
    tract = paste0(state, county, tract),
    perc_hisp = B19001I_001E / B19001_001E,
    perc_over100k = (B19001_014E + B19001_015E + B19001_016E + B19001_017E) 
                     / B19001_001E,
    perc_under100k = (B19001_001E + B19001_002E + B19001_003E + B19001_004E +
                      B19001_005E + B19001_006E + B19001_007E + B19001_008E +
                       B19001_009E + B19001_010E + B19001_011E + B19001_012E + 
                        B19001_013E) / B19001_001E,
    perc_speak_both = (B16006_004E + B16006_005E) / B16006_001E,
    perc_no_both = (B16006_006E + B16006_007E) / B16006_001E
  ) %>% 
  filter(
    !is.na(perc_hisp), 
    !is.na(perc_over100k),
    !is.na(perc_under100k),
    !is.na(perc_speak_both),
    !is.na(perc_no_both)
  )
```

This correlation plot outlines the correlation between households that 
are hispanic, speak Spanish and English well, and have a household income of 
under $100k a year. 

```{r}
library(corrplot)

correlationplot <- bay_multiple_tract %>% 
  select(
    perc_hisp,
    perc_speak_both,
    perc_under100k
  ) %>% 
  cor()

corrplot(
  correlationplot, 
  method = "number",
  type = "upper"
)
```

This correlation plot outlines the correlation between households that 
are hispanic, speak Spanish and do not speak English well, and have a household 
income of under $100k a year. 

```{r}
library(corrplot)

correlationplot <- bay_multiple_tract %>% 
  select(
    perc_hisp,
    perc_no_both,
    perc_under100k
  ) %>% 
  cor()

corrplot(
  correlationplot, 
  method = "number",
  type = "upper"
)
```

This correlation plot outlines the correlation between households that 
are hispanic, speak Spanish and speak English well, and have a household 
income of over $100k a year.

```{r}
library(corrplot)

correlationplot <- bay_multiple_tract %>% 
  select(
    perc_hisp,
    perc_speak_both,
    perc_over100k,
  ) %>% 
  cor()

corrplot(
  correlationplot, 
  method = "number",
  type = "upper"
)
```

This correlation plot outlines the correlation between households that 
are hispanic, speak Spanish and do not speak English well, and have a household 
income of over $100k a year.

```{r}
library(corrplot)

correlationplot <- bay_multiple_tract %>% 
  select(
    perc_hisp,
    perc_no_both,
    perc_over100k
  ) %>% 
  cor()

corrplot(
  correlationplot, 
  method = "number",
  type = "upper"
)
```

The correlation plots show that the percentage of Hispanic people in the Bay 
Area whose households speak Spanish and speak English well and make under $100k 
a year is 55%. Thee percentage of Hispanic people in the Bay Area that speak
Spanish and do not speak English well and have a household income of over $100K 
a year is 32%. An assumption that could come from this data is that the ability
to speak both Spanish and English results in a higher income. 

However, there are many other factors that contribute to income such as 
education levels, access to transportation, household responsibilities, and 
many other factors. It is important to consider other factors and understand
biases despite data availability. 


Sources Used: 

https://stanfordfuturebay.github.io/course/introduction.html

https://www.census.gov/programs-surveys/acs/microdata.html

https://api.census.gov

https://api.census.gov/data/2018/acs/acs5

https://api.census.gov/data/2018/acs/acs5/groups/B19001.html

https://api.census.gov/data/2018/acs/acs5/groups/B16006.html

