---
title: "Assignment 2"
author: "Kimberly Juarez-Rico"
date: "9/26/2020"
output: html_document
---
This assignment focused on performing equity analysis by using Census, American 
Communities Survey (ACS), and Public Use Microdata (PUMS) data. Survey data 
from censusapi comes in "buckets" or different catergories for grouping 
together data on a larger scale. For example, the buckets for different income
levels sometimes range from $30k - $40k. However, PUMS allows for data to be 
analyzed on a smaller scale at the household or even individual level. For
this assignment, ACS 5 year and 1 year data and 5 year PUMS were used to 
analyze different factors that affect education and educational outcomes. 

For this assignment, I chose to focus on Sonoma County when for parts 1 and 3
that specified that the analysis could focus on just one Bay Area county in
particular. Part 2 asked for an analysis on the entire Bay Area.
I chose Sonoma County because I grew up there so I am interested in analyzing 
the data from this area.

Part 1:
The first part of the assignment was to create an analysis on educational 
attainment by race in the county of choice. I used ACS C15002 data which 
groups together educational attainment by race and sex. I had to use 
C15002 data because this was one the only data set that included race
into the analysis of educational attainment. 

First, the necessary packages were installed that would make it easier to 
use the ACS and Census data:

```{r setup, include = F}
knitr::opts_chunk$set(warning = F, message = F, echo = T, eval = T, fig.show = T)
```

```{r}
library(tidyverse)
library(sf)
library(tigris)
library(leaflet)
library(censusapi)
library(tidycensus)
```

Census data was downloaded onto my computer:

```{r}
Sys.setenv(CENSUS_KEY="4f290b276a2a5526b13d917c76884e70b99ba725")

acs_vars_2018_5yr <-
  listCensusMetadata(
    name = "2018/acs/acs5",
    type = "variables"
  )

acs_vars_2018_1yr <-
  listCensusMetadata(
    name = "2018/acs/acs1",
    type = "variables"
  )

acs_vars_2019_1yr <-
  listCensusMetadata(
    name = "2019/acs/acs1",
    type = "variables"
  )

saveRDS(acs_vars_2018_5yr, "acs_vars_2018_5yr")
saveRDS(acs_vars_2018_1yr, "acs_vars_2018_1yr")
saveRDS(acs_vars_2019_1yr, "acs_vars_2019_1yr")
```

This chuck creates a data set that calculates the total number of 
people surveyed for the 15002 data set so that it can then be used to 
manipulate the data set in C15002. 

```{r}
soco_education_total <-
  getCensus(
    name = "acs/acs5",
    vintage = 2018,
    region = "county:097",
    regionin = "state:06",
    vars = c("group(B15002)")
  )%>%
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2018_5yr %>%
      select(name, label),
    by = c("variable" = "name")
  ) %>%
  select(-variable) %>%
  separate(
    label,
    into = c(NA,NA,"education"),
    sep = "!!"
  ) %>%
   filter(!is.na(education))%>%
   summarize(
         total = sum
          (estimate,
            na.rm =T)
   )
soco_education_total
```

Specific race labels were created according to the corresponding data group. 
C15002 is divided by race in that a letter of the alphabet is added to the end
of the title to distinguish between racial groups:

```{r}
census_race_labels <- 
  c(
    "White Alone",
    "Black or African American",
    "American Indian and Alaska Native Alone",
    "Asian Alone",
    "Native Hawaiian and Other Pacific Islander Alone",
    "Some Other Race Alone",
    "Two or More Races",
    "White Alone, Not Hispanic or Latino",
    "Hispanic or Latino"
  )

soco_education_race <-
  1:9 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2018,
      region = "county:097",
      regionin = "state:06",
      vars = paste0("group(C15002",LETTERS[x],")")
    ) %>%
      select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
      pivot_longer(
        ends_with("E"),
        names_to = "variable",
        values_to = "estimate"
      ) %>%
      left_join(
        acs_vars_2018_1yr %>% 
          select(name, label), 
        by = c("variable" = "name")
      ) %>% 
      select(-variable) %>% 
      separate(
        label,
        into = c(NA,NA,"sex"),
        sep = "!!"
      ) %>% 
      filter(!is.na(sex)) %>% 
      mutate(race = census_race_labels[x])%>%
      mutate(soco_education_total,
        percentageOfPop = 
          (estimate / total)*100,
          na.rm = T
      )
  })

soco_education_race
```

This creates a bar graph of educational attainment by race for people living 
in Sonoma County over 25 years old:

```{r}
soco_plot<-
soco_education_race %>% 
  group_by(sex, race) %>% 
  summarize(percentageOfPop = sum(percentageOfPop)) %>% 
  ggplot() +
  geom_bar(
    aes(
      x = race %>% factor(),
      y = percentageOfPop,
      fill = sex
    ),
     stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Educational Attainment",
    y = "Percentage of people (over 25)",
    title = "Sonoma County educational attainment by race",
    fill = "Gender of person"
  ) +
  coord_flip()

soco_plot
```

The percentage of the population that has the highest percentage of 
educational attainment are White and White (non-Hispanic). 

Part 2: Estimating the number and percentage of K-12 students that do not 
have internet access at home using PUMS data. 

Load in the PUMS 2018 data:

```{r}
census_api_key("4f290b276a2a5526b13d917c76884e70b99ba725")

pums_vars_2018 <- 
  pums_variables %>%
  filter(year == 2018, survey == "acs5")

pums_vars_2018_distinct_hh <- 
  pums_vars_2018 %>%
  distinct(var_code, var_label, data_type, level) %>% 
  filter(level == "housing")
```

Categorize PUMS data so that the only data sets presented are the number of 
households with or without access to the internet and 
number of persons under the age of 18 in each household:

```{r}
ca_pums <- get_pums(
  variables = c(
    "PUMA",
    "ACCESS",
    "R18"
  ),
  state = "CA",
  year = 2018,
  survey = "acs5",
  recode = T
)
```

Aggregate the data so that only data from Sonoma County is presented: 

```{r}
ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

soco_name <- "Sonoma"

soco <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME %in% soco_name)

soco_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[soco, ] %>% 
  st_set_geometry(NULL) %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()

soco_pums <-
  ca_pums %>% 
  filter(PUMA %in% soco_pumas$PUMACE10)
```

The final analysis involves finding the total percentage of children who
do not have access to the internet. "Yes, by paying by cell phone company or 
Internet service provider" is the label for households that have internet
access. If a household does not fall under this label and if a person under
18 lives in this household, then the child is counted as not having access
to the internet. The final map shows the distribution of children in 
Sonoma County who do not have access to the internet:

```{r}
soco_internet <-
  soco_pums %>% 
  filter(!duplicated(SERIALNO)) %>% 
  mutate(
    child_nointernet = ifelse(
      (ACCESS_label != "Yes, by
      paying by cell phone company or Internet service provider") &
        (R18 != 0),
      WGTP,
      0
    )
  ) %>% 
  group_by(PUMA) %>% 
  summarize(
    perc_child_nointernet =
      sum(child_nointernet, na.rm =T)/sum(WGTP, na.rm = T)*100
  ) %>% 
  left_join(
    soco_pumas %>% 
      select(PUMACE10),
    by = c("PUMA" = "PUMACE10")
  ) %>% 
  st_as_sf()

pums_pal <- colorNumeric(
  palette = "Blues",
  domain = soco_internet$perc_child_nointernet
)

leaflet() %>%
  addTiles() %>% 
  addPolygons(
    data = soco_internet,
    fillColor = ~pums_pal(perc_child_nointernet),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.5,
    weight = 1,
    label = ~paste0(
      round(perc_child_nointernet), 
      "% of households with no internet and with children under 18"
    ),
    highlightOptions = highlightOptions(
      weight = 2,
      opacity = 1
    )
  ) %>% 
  addLegend(
    data = soco_internet,
    pal = pums_pal,
    values = ~perc_child_nointernet,
    title = "% households<br>with no internet<br> with children under 18"
  )
```

The map shows that there are some differences in internet access within 
different communities in Sonoma County. There is a 7% difference in the 
internet access in the "outer" parts of the county. Santa Rosa is the biggest 
city in the county by population. According to DataUSA, Windsor's average 
houshold income is $99,792 while Santa Rosa's average household income is 
$71,374. The difference in income could help explain why there are 7% less 
households with internet access in Santa Rosa versus areas more north such 
as Windsor.

Part 3: Analysis of educational mobility in Sonoma County. 

By using 2018 and 2019 educational attainment and mobility data, an analysis
of educational mobility for Sonoma County can be conducted. 

Useful ACS Codes: 

Code: B07009
Description:"Geographical mobility in the past year by educational 
attainment for current residence in the United States"
What it does: Counts current population in a given year, the combination of 
people who have remained and people who have immigrated in.

Code: B07409
Description: "Geographical Mobility in the past year by educational attainment 
for residence 1 year ago in the United States"
What it does: Counts people who have remained and people who were here a year 
ago but have emigrated elsewhere.


Counts the population of educational attainment beyond high school in 2019 that 
includes both people who have remained in Sonoma County and people who have 
immigrated into Sonoma County: 

```{r}
soco_edmobility_current_19 <- 
  getCensus(
    name = "acs/acs1",
    vintage = 2019,
    region = "county:097",
    regionin = "state:06",
    vars = c("group(B07009)")
  ) %>% 
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_1yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable)%>%
  separate(
    label,
    into = c(NA,NA,"mobility","education"),
    sep = "!!"
  )%>%
  mutate(
    mobility = ifelse(
      mobility %in% c("Same house 1 year ago:", "Moved within same county:"),
      "Here last year",
      "Inflow"
    )
  )%>%
  filter(!is.na(education)) %>%
  group_by(mobility, education) %>%
  summarize(estimate = sum(estimate))

soco_edmobility_current_19
```

Counts the population of educational attainment beyond high school in 2019 that 
includes both people who have remained in Sonoma COunty and people who have
emigrated to places out of Sonoma County:

```{r}
soco_edmobility_lastyear_19 <- 
  getCensus(
    name = "acs/acs1",
    vintage = 2019,
    region = "county:097",
    regionin = "state:06",
    vars = c("group(B07409)")
  ) %>% 
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_1yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable)%>%
  separate(
    label,
    into = c(NA,NA,"mobility","education"),
    sep = "!!"
  )%>%
  mutate(
    mobility = ifelse(
      mobility %in% c("Same house 1 year ago:", "Moved within same county:"),
      "Outflow",
      "Here last year"
    )
  )%>%
  filter(!is.na(education)) %>%
  group_by(mobility, education) %>%
  summarize(estimate = sum(estimate))

soco_edmobility_lastyear_19
```

Counts the population of educational attainment in 2018 that includes both
people who have remained into Sonoma County and people who have immigrated 
in Sonoma County: 

```{r}
soco_edmobility_current_18 <- 
  getCensus(
    name = "acs/acs1",
    vintage = 2018,
    region = "county:097",
    regionin = "state:06",
    vars = c("group(B07009)")
  ) %>% 
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2018_1yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable)%>%
  separate(
    label,
    into = c(NA,NA,"mobility","education"),
    sep = "!!"
  )%>%
  mutate(
    mobility = "Here last year"
  )%>%
  filter(!is.na(education)) %>%
  group_by(mobility, education) %>%
  summarize(estimate = sum(estimate))

soco_edmobility_current_18
```

Takes the population of educational attainment for 2018 and 2019 to calculate
the number of people that have immigrated out of Sonoma County and into 
Sonoma County:

```{r}
soco_flows_19 <-
  rbind(
    soco_edmobility_current_18,
    soco_edmobility_lastyear_19 %>% 
      filter(mobility == "Outflow"),
    soco_edmobility_current_19 %>% 
      filter(mobility == "Inflow"),
    soco_edmobility_current_19 %>% 
      group_by(education) %>% 
      summarize(estimate = sum(estimate)) %>% 
      mutate(mobility = "Here this year")
  ) %>% 
  pivot_wider(
    names_from = mobility,
    values_from = estimate
  ) %>% 
  mutate(
    `External net` = Inflow - Outflow,
    `Internal net` = `Here this year` - `Here last year` - `External net`,
  ) %>% 
  select(
    `Education level` = education, 
    `Internal net`,
    `External net`,
    `Here last year`, 
    `Here this year`, 
    Outflow, 
    Inflow
  )

soco_flows_19
```

The table shows that there are more people with Bachelor's degrees and with high 
school equivalent degrees moving into the county. For the other groups however,
the external net is higher meaning that more people are moving out of the county
for these groups. This could be because Sonoma County's main industry is 
wine and viticulture where most jobs require high school degrees and Bachelor's
degrees. 

Sources:

https://datausa.io/profile/geo/santa-rosa-ca/#:~:text=In%202018%
2C%20Santa%20Rosa%2C%20CA,median%20household%20income%20of%20%2471%2C347
.&text=The%205%20largest%20ethnic%20groups,%2DHispanic)%20(3.61%25).

https://stanfordfuturebay.github.io/course/introduction.html

https://www.census.gov/programs-surveys/acs/microdata.html

https://api.census.gov

https://api.census.gov/data/2018/acs/acs5

https://api.census.gov/data/2018/acs/acs1

https://api.census.gov/data/2019/acs/acs1

