---
title: "Assignment 7"
author: "Kimberly Juarez-Rico"
date: "11/25/2020"
output: html_document
---

This assignment involves the use of emission data and parcel data. 

The first part of the assignment focused on vechicles emissions in Sebastopol, 
CA and the second part of the assignment is focused on parcel data emissions
in Sebastopol, cA.

```{r setup, include = F}
knitr::opts_chunk$set(warning = F, message = F, echo = T, eval = F)
```

Loaded necessary libraries and JSON Data:

```{r}
library(tigris)
library(tidyverse)
library(censusapi)
library(sf)
library(leaflet)
library(mapboxapi)
library(jsonlite)

Sys.setenv(CENSUS_KEY="4f290b276a2a5526b13d917c76884e70b99ba725")

acs_vars_2018_5yr <-
  listCensusMetadata(
    name = "2018/acs/acs5",
    type = "variables"
  )

june_patterns <- read_csv("neighborhood_patterns.gz")
```

Loaded CA census block groups and then filtered them to Sebastopol, CA:

```{r}
ca_cbgs <- block_groups("CA", cb = T, progress_bar = F)

seb_boundary <- places("CA", cb = T, progress_bar = F) %>% 
  filter(NAME == "Sebastopol")

seb_cbgs <- 
  ca_cbgs %>% 
  st_centroid() %>% 
  .[seb_boundary, ] %>% 
  st_set_geometry(NULL) %>% 
  left_join(ca_cbgs %>% select(GEOID)) %>% 
  st_as_sf()
```

Filtered Sebastopol commute patterns by using the June commute patterns data:

```{r}
seb_patterns <-
  june_patterns %>% 
  filter(area %in% seb_cbgs$GEOID)

rm(june_patterns)
```

Creates a data set that filters the devices in Sebastopol:

```{r}
seb_json <-
  seb_patterns$device_home_areas[1] %>% 
  fromJSON() %>%
  unlist() %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  rename(
    origin_cbg = "rowname",
    device_count = "."
  )

head(seb_json)
```

Counts the devices that have an origin in Sebastopol;
```{r}
seb_origin_cbgs <-
  1:nrow(seb_patterns) %>% 
  map_dfr(function(row){
    seb_json <-
      seb_patterns$device_home_areas[row] %>% 
      fromJSON() %>%
      unlist() %>% 
      as.data.frame() %>% 
      rownames_to_column() %>% 
      rename(
        seb_origin_cbg = "rowname",
        device_count = "."
      ) %>% 
      mutate(
        device_count = ifelse(
          device_count == 4,
          3,
          device_count
        )
      )
    
  })
```

Counts the total number of devices:
```{r}
seb_origin_cbgs_merged <-
  seb_origin_cbgs %>% 
  group_by(seb_origin_cbg) %>% 
  summarize(
    device_count = max(device_count)
  )
```

Reads the home_panel_summary:
```{r}
seb_home_panel_summary <- read_csv("home_panel_summary.gz")
```

Filters CA census block groups:

```{r}
ca_cbgs_pop <-
  counties("CA", cb = T, progress_bar = F) %>%
  pull(COUNTYFP) %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2018,
      region = "block group:*",
      regionin = paste0("state:06+county:", x),
      vars = "B01001_001E"
    )
  }) %>% 
  transmute(
    census_block_group =
      paste0(state,county,tract,block_group),
    pop = B01001_001E
  )
```

Filters census block group data for Sebastopol:

```{r}
seb_origin_cbgs_pop <-
  seb_origin_cbgs_merged %>% 
  left_join(
    ca_cbgs_pop,
    by = c("seb_origin_cbg" = "census_block_group")
  )

sum(is.na(seb_origin_cbgs_pop$pop))
```

Counts the number of devices residing in Sebastopol:

```{r}
seb_origin_cbgs_normalized <-
  seb_origin_cbgs_pop %>% 
  filter(!is.na(pop)) %>% 
  left_join(
    seb_home_panel_summary %>% 
      select(seb_origin_cbg = census_block_group, number_devices_residing)
  ) %>% 
  mutate(
    visits = (device_count * pop / number_devices_residing) %>% round()
  ) %>% 
  left_join(ca_cbgs %>% select(seb_origin_cbg = GEOID)) %>% 
  st_as_sf()
```

Creates a map of the routes to Sebastopol by using devices:

```{r}
visits_pal <- colorNumeric(
  palette = "Reds",
  domain = seb_origin_cbgs_normalized %>% 
    arrange(desc(visits)) %>% 
    pull(visits) %>% 
    .[-c(1:6)]
)

leaflet() %>% 
  addMapboxTiles(
    style_id = "light-v9",
    username = "mapbox"
  ) %>% 
  addPolygons(
    data = seb_cbgs,
    fill = F
  ) %>% 
  addPolygons(
    data = seb_origin_cbgs_normalized,
    fillColor = ~visits_pal(visits),
    color = "red",
    weight = 1,
    fillOpacity = 0.75,
    label = ~visits
  )
```

Creates the origin and destination points of trips to Sebastopol:

```{r}
seb_origin <-
  seb_origin_cbgs_normalized %>% 
  st_centroid() %>% 
  st_coordinates()

seb_destination <-
  seb_boundary %>% 
  st_centroid() %>% 
  st_coordinates()

seb_route <- 
  1:nrow(seb_origin) %>%
  map_dfr(function(x){
    mb_directions(
      origin = seb_origin[x, ],
      destination = seb_destination,
      profile = "driving-traffic"
    )
  }) %>% 
  st_as_sf()
```

Creates a map of routes to Sebastopol:

```{r}
leaflet() %>%
  addMapboxTiles(
    style_id = "light-v9",
    username = "mapbox"
  ) %>%
  addPolylines(
    data = seb_route
  )
```

Creates a data set of Sebastopol travel times by using ACS 5 year data for 
the duration of trips to Sebastopol. The duration of trips will then determine
the emissions produced:

```{r}
seb_travel_time_mode <-
  counties("CA", cb = T, progress_bar = F) %>%
  pull(COUNTYFP) %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2018,
      region = "block group:*",
      regionin = paste0("state:06+county:", x),
      vars = "group(B08134)"
    )
  }) %>% 
  mutate(
    cbg =
      paste0(state,county,tract,block_group)
  ) %>%
  filter(cbg %in% seb_origin_cbgs_normalized$seb_origin_cbg) %>% 
  select(!c(GEO_ID,state,county,tract,block_group,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2018_5yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable) %>% 
  separate(
    label,
    into = c(NA, NA, "total", "mode", "carpool", "time"),
    sep = "!!"
  ) %>% 
  mutate(
    mode = case_when(
      total %in% c(
        "Less than 10 minutes",
        "10 to 14 minutes",
        "15 to 19 minutes",
        "20 to 24 minutes",
        "25 to 29 minutes",
        "30 to 34 minutes",
        "35 to 44 minutes",
        "45 to 59 minutes",
        "60 or more minutes"
      ) ~ "Total",
      mode == "Drove alone" ~ mode,
      carpool %in% c(
        "In 2-person carpool",
        "In 3-or-more-person carpool"
      ) ~ carpool
    ),
    time = case_when(
      mode == "Total" ~ total,
      mode == "Drove alone" ~ carpool,
      mode == carpool ~ time
    )
  ) %>% 
  filter(!is.na(time)) %>% 
  select(-total, -carpool) %>% 
  pivot_wider(
    names_from = mode,
    values_from = estimate
  ) %>% 
  mutate(
    perc_veh1 = `Drove alone`/Total,
    perc_veh2 = `In 2-person carpool`/Total,
    perc_veh3 = `In 3-or-more-person carpool`/Total
  )
```

The number of trips are then combined with the origin and destination data
provided by the Sebastopol commute patterns:

```{r}
seb_trips <-
  seb_origin_cbgs_normalized %>% 
  cbind(
    seb_route %>% 
      st_set_geometry(NULL)
  ) %>% 
  mutate(
    time = case_when(
      duration < 10 ~ "Less than 10 minutes",
      duration < 15 ~ "10 to 14 minutes",
      duration < 20 ~ "15 to 19 minutes",
      duration < 25 ~ "20 to 24 minutes",
      duration < 30 ~ "25 to 29 minutes",
      duration < 35 ~ "30 to 34 minutes",
      duration < 45 ~ "35 to 44 minutes",
      duration < 60 ~ "45 to 59 minutes",
      TRUE ~ "60 or more minutes"
    )
  ) %>% 
  left_join(
    seb_travel_time_mode %>% 
      select(
        seb_origin_cbg = cbg,
        time,
        perc_veh1,
        perc_veh2,
        perc_veh3
      ),
    by = c("seb_origin_cbg", "time")
  ) %>% 
  mutate(
    vehicles = 
      visits * perc_veh1 + 
      visits * perc_veh2 / 2 +
      visits * perc_veh3 / 3,
    vmt = vehicles * distance * 2
  )
```

Sums the total number of trips by counting the number of devices:

```{r}
recorded_visits <- sum(seb_trips$device_count)/
  sum(seb_patterns$raw_device_counts)

recorded_visits
```

Outputs the total vehicle miles traveled by the number of trips:

```{r}
sum(seb_trips$vmt, na.rm = T)/recorded_visits
```

Executes 2017 vehicle EMFAC data:

```{r}
emfac <- 
  read_csv("EMFAC2017-EI-2011Class-BayAreaAQMD-2020-Summer-20201125144541.csv", 
           skip = 8) %>% 
  transmute(
    Category = `Vehicle Category`,
    Fuel_Type = Fuel,
    Percent_Trips = Trips/sum(Trips),
    Percent_Miles = VMT/sum(VMT),
    `gCO2_Running_Exhaust` = CO2_RUNEX,
    `gCO2_Start_Exhaust` = CO2_STREX
  )

emfac
```

Calculates the emissions given by commuters to Sebastopol, CA:

```{r}
seb_trips_ghg <-
  emfac %>% 
  mutate(
    trips = Percent_Trips * sum(seb_trips$visits, na.rm = T),
    vmt = Percent_Miles * sum(seb_trips$vmt, na.rm = T),
    ghg = vmt*gCO2_Running_Exhaust + trips*gCO2_Start_Exhaust*2
  )

sum(seb_trips_ghg$ghg)*1e-6/recorded_visits
```


Part 2: Building Emissions, ZIPCODE 95472 for Sebastopol

The second part of the assignment focused on calculating emissions based on the 
energy use by parcel for Sebastopol, CA. 

Inputs PG&E data for the summer of 2020:

```{r}
pge_data <- 
  2020 %>% 
  map_dfr(function(yr){
    
    factor <- 
      pge_elec_emissions_factor %>% 
      filter(year == yr) %>% 
      pull(factor)
    
    1:2 %>% 
      map_dfr(function(quarter){
        
        c("Electric","Gas") %>% 
          map_dfr(function(type){
            
            filename <- 
              paste0(
                "PGE_",
                yr,
                "_Q",
                quarter,
                "_",
                type,
                "UsageByZip.csv"
              )
            
            temp <- read_csv(filename)
            
            temp <-
              temp %>% 
              rename_all(toupper) %>% 
              mutate(
                TOTALKBTU = ifelse(
                  substr(CUSTOMERCLASS,1,1) == "E",
                  TOTALKWH * 3.412,
                  TOTALTHM * 99.976
                ),
                TOTALTCO2E = ifelse(
                  substr(CUSTOMERCLASS,1,1) == "E",
                  TOTALKWH/1000 * factor * 0.000453592,
                  TOTALTHM * 0.00531
                )
              ) %>% 
              select(
                ZIPCODE,
                YEAR,
                MONTH,
                CUSTOMERCLASS,
                TOTALKBTU,
                TOTALTCO2E,
                TOTALCUSTOMERS
              )
            
          })
        
      })
    
  })
```

Filters ZCTAS data for the US, then Sonoma County, and then Sebastopol:

```{r}
us_zips <- 
  zctas(cb = T, progress_bar = F)

sonoma_zips <- 
  us_zips %>% 
  st_centroid() %>% 
  .[counties("CA", cb = T, progress_bar = F) %>% filter(NAME == "Sonoma"), ] %>% 
  st_set_geometry(NULL) %>% 
  left_join(us_zips %>% select(GEOID10)) %>% 
  st_as_sf()

seb_zips<-
  sonoma_zips %>%
  filter(ZCTA5CE10 == 95472)
```

Filters Sebastopol PG&E data from the total PG&E data set to electric and gas
residential use and commercial and gas residential use:

```{r}
seb_pge_data <-
  pge_data %>% 
  filter(ZIPCODE %in% seb_zips$ZCTA5CE10) %>% 
  filter(CUSTOMERCLASS %in% c(
    "Elec- Commercial",
    "Elec- Residential",
    "Gas- Commercial",
    "Gas- Residential"
  )) %>% 
  mutate(
    UseCodType = substr(CUSTOMERCLASS,0,20)
  ) %>%
  group_by(ZIPCODE, UseCodType, MONTH, YEAR) %>% 
  summarize(
    TOTALKBTU = sum(TOTALKBTU, na.rm=T),
    TOTALTCO2E = sum(TOTALTCO2E, na.rm=T), 
    TOTALCUSTOMERS = mean(TOTALCUSTOMERS, na.rm=T)
  ) %>% 
  group_by(ZIPCODE,UseCodType, MONTH) %>%
  summarize_at(
    vars(TOTALKBTU,TOTALTCO2E,TOTALCUSTOMERS),
    sum,
    na.rm=T
  )
```

Filters energy data for just the month of June 2020:

```{r}
seb_june_2020 <-
  seb_pge_data %>%
  filter(MONTH == 6)
```

Summarizes energy use by KBTUs and CO2E emissions:

```{r}
seb_pge_june_2020 <-
  seb_june_2020%>%
group_by(UseCodType,TOTALKBTU, TOTALTCO2E)%>%
 summarize(
  TOTALKBTU = sum(TOTALKBTU, na.rm = T),
  TOTALTCO2E = sum(TOTALTCO2E, na.rm = T)
  )
```

Creates a map of June 2020 energy use in Sebastopol:

```{r}
ggplot(
  seb_june_2020, 
  aes(
    x = as.factor(MONTH), 
    y = TOTALKBTU/1000000
  )
) + 
  geom_bar(stat = "identity", aes(fill = ENERGYTYPE), position = "dodge") + 
  labs(x = "Month", y = "GBTU", title = "Sebastopol Energy Usage, June 2020") + 
  scale_fill_discrete(name="Energy Type",labels = c("Electricity","Gas"))
```

Creates a map of June 2020 emissions in Sebastopol:

```{r}
ggplot(
  seb_june_2020, 
  aes(
    x = as.factor(MONTH), 
    y = TOTALTCO2E
  )
) + 
  geom_bar(stat = "identity", aes(fill = ENERGYTYPE), position = "dodge") + 
  labs(x = "Month", y = "tCO2e", title = "Sebastopol Emissions, June 2020") + 
  scale_fill_discrete(name="Energy Type",labels = c("Electricity","Gas"))
```


I was able to find parcel data for Sonoma County by using the County of
Sonoma website that has open source parcel data for direct download.
This is probably due to the fact that the area is prone to wildfires so openly 
available parcel data may be requested often. 

Bring in parcel data:
```{r}
soco_parcel_data <- read_csv("Sonoma_County_Parcels.csv")
```

Filter to just Sebastopol by city name, building type, and 95472 zip code:
```{r}
seb_parcel_filter <-
  soco_parcel_data%>%
  select(
    OBJECTID, UseCodType,TRADesc, LndSizAcre, LndSizSqft, MailCtySta, MailZip5,
    SitusFrm1
  )%>%
  filter(
    TRADesc %in% c("Sebastopol"),
    
    UseCodType %in%
      c(
        "Residential",
        "Commercial"),
    
    MailZip5 %in% c("95472")
  )%>%
```

Filters the parcel data to square feet:

```{r}
seb_parcel_final <-
seb_parcel_filter%>%
summarize(
  TotalSqFt = sum(LndSizSqft, na.rm = T))
```

Total commercial electricity use per parcel square feet in Sebastopol for the 
month of June 2020:

```{r}
elec_com_use_per_sqft = seb_pge_june_2020[1,2] / seb_parcel_final[1,2]
```

Total commercial emissions per parcel square feet in Sebastopol for the month 
of June 2020:

```{r}
elec_com_emissions_per_sqft = seb_pge_june_2020[1,3] / seb_parcel_final[1,2]
```

Total residential electricity use per parcel square feet in Sebastopol for the 
month of June 2020:

```{r}
elec_res_use_per_sqft = seb_pge_june_2020[2,2] / seb_parcel_final[2,2]
```

Total residential electricity emissions per parcel square feet in Sebastopol for 
the month of June 2020:

```{r}
elec_res_emissions_per_sqft = seb_pge_june_2020[2,3] / seb_parcel_final[2,2]
```

Total commercial gas use per parcel square feet in Sebastopol for the month of 
June 2020:

```{r}
gas_com_use_per_sqft = seb_pge_june_2020[3,2] / seb_parcel_final[1,2]
```

Total gas emissions per parcel square feet in Sebastopol for the month of 
June 2020: 

```{r}
gas_com_emissions_per_sqft = seb_pge_june_2020[3,3] / seb_parcel_final[1,2]
```

Total residential gas use per parcel square feet in Sebastopol for the month of 
June 2020:

```{r}
gas_elec_use_per_sqft = seb_pge_june_2020[4,2] / seb_parcel_final[2,2]
```

Total residential gas emissions per parcel square feet in Sebastopol for the 
month of June 2020:

```{r}
gas_elec_emissions_per_sqft = seb_pge_june_2020[4,3] / seb_parcel_final[2,2]
```

Sources Used:

https://stanfordfuturebay.github.io/course/introduction.html

https://sonomacounty.ca.gov/PRMD/Administration/GIS/Geographical-Information-Systems/




