---
title: "Assignment 1"
author: "Kimberly Juarez-Rico"
date: "9/18/2020"
output: html_document
---

The following assignment demonstrates how to read and manipulate energy
data files in order to create bar plots and geospatial maps.

PGE puts out Excel files of energy use data for each quarter that anyone
can download and analyze. I downloaded the 2017 - 2020 data from their site
(with 2020 only having Q1 and Q2 so far). 

I then restricted zip codes to the Bay Area in order to analyze
energy usage in the Bay Area from 2017 to 2020 and changes in residential
electric energy use after the pandemic began in Q2 of 2020. 

Part 1: This part of the assignment creates a bar plot of residential and
commercial electric and gas energy data for Bay Area zip codes from
quarter 1 of 2017 to quarter 2 of 2020. At the time of analysis, data from
quarters 3 and 4 of 2020 were not yet available so they were not included,
but could potentially be in the future. 



```{r setup, include = F}
knitr::opts_chunk$set(warning = F, message = F, echo = F, eval = T)
```

This imports the libraries needed:

```{r}
library(tidyverse)
library(sf)
library(tigris)
library(leaflet)
library(plotly)
library(gtools)
library(dplyr)
```

Creates a data set for all of the electricity use data from Q1 2017 to Q2
2020:

```{r}
pge_elec <- NULL
years <- 2017:2020
quarters <- 1:4
type <- "Electric"

for (year in years){
  for(quarter in quarters) {
      
  if ( (year == 2020) & ((quarter == 3) | (quarter == 4))){
    next
  }
  
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  
  temp <- read_csv(filename)
  
 pge_elec <- rbind(pge_elec, temp)
  }
}
```

Filters the electricity data so that the data frame only contains total
energy usage and converts the energy data values from kWh to kBTU.
The energy is converted to kBTU so that it can also be compared with gas
data which is usually in therms. Therms will also be converted to kBTU:

```{r}
elec_change <-
  pge_elec %>% 
  filter(
    CUSTOMERCLASS %in% 
      c(
        "Elec- Residential",
        "Elec- Commercial"
      )
  )%>% 
  select(
    !c(COMBINED, AVERAGEKWH,TOTALCUSTOMERS)
  ) %>% 
  group_by(MONTH, CUSTOMERCLASS) %>% 
   mutate (
    TOTALKBTU =
      TOTALKWH * 3412.14
    )%>%
   select(!TOTALKWH)

elec_change
```

This creates a data set for all of the gas use data from Q1 2017 to Q2
2020:

```{r}
pge_gas <- NULL
years <- 2017:2020
quarters <- 1:4
type <- "Gas"

for (year in years){
  for(quarter in quarters) {
      
  if ( (year == 2020) & ((quarter == 3) | (quarter == 4))){
    next
  }
  
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  
 pge_gas <- rbind(pge_gas, temp)
  }
}
```

This filters the gas data so that the data frame only contains total
energy usage and converts from therms to kBTU:

```{r}
gas_change <-
  pge_gas %>% 
  filter(
    CUSTOMERCLASS %in% 
      c(
        "Gas- Residential",
        "Gas- Commercial"
      )
  )%>% 
  select(
    !c(COMBINED, AVERAGETHM,TOTALCUSTOMERS)
  ) %>% 
  group_by(MONTH, CUSTOMERCLASS) %>% 
   mutate (
    TOTALKBTU =
      TOTALTHM * 99976.1
    )%>%
   select(!TOTALTHM)

gas_change
```

This binds the total electricity use and gas use data into a dataframe that will
then be plotted: 

```{r}
pge_elec_gas <- NULL
  pge_elec_gas <- rbind(pge_elec_gas, elec_change, gas_change)
```

This filters out zip code data for the Bay Area:

```{r}
usa_zips <- 
  zctas(cb = T, progress_bar = F)

ca_counties <- counties("CA", cb = T, progress_bar = F)

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_counties <-
  ca_counties %>%
  filter(NAME %in% bay_county_names)

bay_zips <-
  usa_zips %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_set_geometry(NULL) %>% 
  left_join(usa_zips %>% select(GEOID10)) %>% 
  st_as_sf()

pge_elec_gas_bay <-
  pge_elec_gas %>% 
  filter(CUSTOMERCLASS %in% 
      c(
        "Elec- Residential",
        "Elec- Commercial",
        "Gas- Residential",
        "Gas- Commercial"
      )
  ) %>% 
  mutate(
    ZIPCODE = ZIPCODE %>% as.character()
  ) %>% 
  group_by(ZIPCODE, MONTH, YEAR, CUSTOMERCLASS) %>% 
  summarize(
    TOTALKBTU = sum(TOTALKBTU, na.rm =T)
  ) %>%
  right_join(
    bay_zips %>% select(GEOID10),
    by = c("ZIPCODE" = "GEOID10")
  ) %>% 
  st_as_sf() %>% 
  st_transform(4326)%>%
  select(!geometry)

pge_elec_gas_bay
```

Filters the energy data of the Bay Area to just the factors needed 
for the analysis: year, month, and customerclass:

```{r}
pge_total <-
  pge_elec_gas_bay %>% 
  filter(
    CUSTOMERCLASS %in% 
      c(
        "Elec- Residential",
        "Elec- Commercial",
        "Gas- Residential",
        "Gas- Commercial"
      )
  ) %>% 
  group_by(YEAR,MONTH,CUSTOMERCLASS) %>% 
  summarize(
    TOTALKBTU = 
      sum(
        TOTALKBTU, 
        na.rm = T
      )
  )

pge_total
```

Combines YEAR and MONTH into one column so that the information is more
clearly summarized in the plot:

```{r}
  pge_total$dates <-
  paste0(pge_total$YEAR, "-", pge_total$MONTH)
```

Further filters the energy data into date and month, customer class, and total
kBTU:

```{r}
pge_final <-
  pge_total %>% 
  select(dates,CUSTOMERCLASS,TOTALKBTU)

pge_final
```

This creates a plot of the data:

```{r}
pge_chart <-
  pge_final%>%
  ggplot() +
  geom_bar(
    aes(
      x = dates %>% factor(),
      y = TOTALKBTU,
      fill = CUSTOMERCLASS
    ),
    stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Year, Month",
    y = "kBTU",
    title = "PG&E Territory Monthly Energy Usage",
    fill = "Type"
  )+ 
  theme(axis.text.x=element_text(angle=50, size=5, vjust=0.5))

pge_chart

pge_chart %>% ggplotly()
```

Part 2: This part of the assignment analyzes residential energy usage
for Bay Area block groups before and after the pandemic began. The pandemic
technically began in quarter 1 of 2020, but quarter 2 of 2020 are where the
changes more most likely seen because shelter-in-place began late in
quarter 1 of 2020. Therefore, comparing the same quarter from the year before,
quarter 2 of 2019 to quarter 2 of 2020 would most likely show a difference
in residential electric energy usage. Comparing the same quarter but in 
different years can also help decrease the chance that changes in residential
electric energy usage are not just due to changes of the seasons.

Imports the CSV file of Q2 2019, before pandemic began:

```{r}
pge_19_Q2_elec <- read_csv("PGE_2019_Q2_ElectricUsageByZip.csv")
  pge_19_Q2_elec <- rbind(pge_19_Q2_elec) 
```

Filters out the residential electricity data and converts to kBTU:

```{r}
pge_19_Q2_elec_filters <-
  pge_19_Q2_elec %>%
  filter(
    CUSTOMERCLASS %in% 
      ("Elec- Residential")
  )%>% 
  select(
    !c(COMBINED)
  ) %>% 
  group_by(ZIPCODE, MONTH) %>%
   mutate (
     AVERAGEKWH =
      TOTALKWH/TOTALCUSTOMERS,
     na.rm = T,
     
    AVERAGEKBTU_19 =
      AVERAGEKWH * 3412.14
   )%>%
  group_by(ZIPCODE) %>% 
  summarise(AVERAGEKBTU_19 = sum(AVERAGEKBTU_19))

pge_19_Q2_elec_filters
  
```

Imports the CSV file of Q2 2020, after the pandemic began:

```{r}
pge_20_Q2_elec <- read_csv("PGE_2020_Q2_ElectricUsageByZip.csv")
  pge_20_Q2_elec <- rbind(pge_20_Q2_elec) 
```

Filters out the residential electricity data and converts to kBTU:

```{r}
pge_20_Q2_elec_filters <-
  pge_20_Q2_elec %>%
  filter(
    CUSTOMERCLASS %in% 
      ("Elec- Residential")
  )%>% 
  select(
    !c(COMBINED)
  ) %>% 
  group_by(ZIPCODE, MONTH) %>%
   mutate (
     AVERAGEKWH =
      TOTALKWH/TOTALCUSTOMERS,
     
    AVERAGEKBTU_20 =
      AVERAGEKWH * 3412.14
   )%>%
  group_by(ZIPCODE) %>% 
  summarise(AVERAGEKBTU_20 = sum(AVERAGEKBTU_20))

pge_20_Q2_elec_filters
```

Filters the data so that only energy data of Bay Area block groups is 
included:

```{r}
usa_zips <- 
  zctas(cb = T, progress_bar = F)

bay_zips <-
  usa_zips %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_set_geometry(NULL) %>% 
  left_join(usa_zips %>% select(GEOID10)) %>% 
  st_as_sf()

bay_cbgs <- block_groups("CA", bay_county_names[1:9], cb = T, progress_bar = F)

bay_cbgs <- 
  bay_county_names %>% 
  map_dfr(function(county) {
    block_groups("CA", county, cb = T, progress_bar = F)
  })

pge_19_Q2_elec_final <-
  pge_19_Q2_elec_filters %>% 
  mutate(
    ZIPCODE = ZIPCODE %>% as.character()
  ) %>% 
  right_join(
    bay_zips %>% select(GEOID10),
    by = c("ZIPCODE" = "GEOID10")
  ) %>% 
  st_as_sf() %>% 
  st_transform(4326)

pge_20_Q2_elec_final <-
  pge_20_Q2_elec_filters %>% 
  mutate(
    ZIPCODE = ZIPCODE %>% as.character()
  ) %>% 
  right_join(
    bay_zips %>% select(GEOID10),
    by = c("ZIPCODE" = "GEOID10")
  ) %>% 
  st_as_sf() %>% 
  st_transform(4326)
```

Creates a data frame that has energy data for the second quarter of 2020
```{r}
pge_Q2_19_20 <-
  pge_19_Q2_elec_final%>%
  st_join(
    pge_20_Q2_elec_final %>% select(AVERAGEKBTU_20),
    by = c("AVERAGEKBTU_TOTAL" = "AVERAGEKBTU_20")
  )

pge_Q2_19_20
```

Creates a new column that takes the difference in values of kBTU
energy usage of quater 2 of 2020 from quarter 2 of 2020. This is possible 
because the same zip codes are used for both columns of data:

```{r}
pge_Q2_19_20$AVERAGEKBTU_Diff <-
  (pge_Q2_19_20$AVERAGEKBTU_20 - pge_Q2_19_20$AVERAGEKBTU_19)
```

Creates an color coded map of energy usage for quarter 2 of 2019. The more 
intense red color indicates greater residential electric energy usage 
for that block group:

```{r}
res_pal <- colorNumeric(
  palette = "Reds",
  domain = pge_Q2_19_20$AVERAGEKBTU_19
)

leaflet() %>% 
  addTiles() %>% 
  addPolygons(
    data = pge_Q2_19_20,
    fillColor = ~res_pal(AVERAGEKBTU_19),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.5,
    weight = 1,
    label = ~paste0(
      round(AVERAGEKBTU_19), 
      " kBTU average in ",
      ZIPCODE
    ),
    highlightOptions = highlightOptions(
      weight = 2,
      opacity = 1
    )
  ) %>% 
  addLegend(
    data = pge_Q2_19_20,
    pal = res_pal,
    values = ~AVERAGEKBTU_19,
    title = "Total Residential<br>kBTU, Q2, 2019"
  )
```

Creates an color coded map of energy usage for quarter 2 of 2020. The more 
intense blue color indicates greater residential electric energy usage 
for that block group:

```{r}
res_pal <- colorNumeric(
  palette = "Blues",
  domain = pge_Q2_19_20$AVERAGEKBTU_20
)

leaflet() %>% 
  addTiles() %>% 
  addPolygons(
    data = pge_Q2_19_20,
    fillColor = ~res_pal(AVERAGEKBTU_20),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.5,
    weight = 1,
    label = ~paste0(
      round(AVERAGEKBTU_20), 
      " kBTU average in ",
      ZIPCODE
    ),
    highlightOptions = highlightOptions(
      weight = 2,
      opacity = 1
    )
  ) %>% 
  addLegend(
    data = pge_Q2_19_20,
    pal = res_pal,
    values = ~AVERAGEKBTU_20,
    title = "Total Residential<br>kBTU, Q2, 2020"
  )
```

Creates an color coded map of the difference in energy usage for block groups. 
The more intense green color indicates a greater change residential electric 
energy usage for that block group from quarter 2 of 2019 to quarter 2 of 2020,
or change from before and after the pandemic:
```{r}
res_pal <- colorNumeric(
  palette = "Greens",
  domain = pge_Q2_19_20$AVERAGEKBTU_Diff
)

leaflet() %>% 
  addTiles() %>% 
  addPolygons(
    data = pge_Q2_19_20,
    fillColor = ~res_pal(AVERAGEKBTU_Diff),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.5,
    weight = 1,
    label = ~paste0(
      round(AVERAGEKBTU_Diff), 
      " kBTU average in ",
      ZIPCODE
    ),
    highlightOptions = highlightOptions(
      weight = 2,
      opacity = 1
    )
  ) %>% 
  addLegend(
    data = pge_Q2_19_20,
    pal = res_pal,
    values = ~AVERAGEKBTU_Diff,
    title = "Difference in <br>Total Residential kBTU <br>
    in Q2 2019 <br> to Q2 2020"
  )
```

Sources: 
https://pge-energydatarequest.com/public_datasets
https://stanfordfuturebay.github.io/course/introduction.html

